MCMC convergence diagnostics

### 2023-06-02 src/docs_paper/appendix.Rmd
Add MCMC convergence asssessment to appendix #48

# Simplified Naomi model description \label{sec:math}

@@

# \textsc{C++} `TMB` user template

@@

This section contains C++ `TMB` code for the negative log-posterior of the simplified Naomi model
For ease of understanding, Table \ref{tab:tmb-lookup} provides correspondence between the mathematical notation used in Section \ref{sec:math} and the variable names used in the `TMB` code, for all hyperparameters and latent field parameters.
For further reference on the `TMB` software see `https://kaskr.github.io/adcomp/_book`, or on the method see @kristensen2016tmb.

```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines("naomi_simple.cpp")}
```

@@

We assessed MCMC convergence and suitability using a range of graphical and numerical tests. These included the potential scale reduction factor $\hat R$, bulk and tail effective sample size (ESS), autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments.

For the time being, this analysis is available from `athowes.github.io/elgm-inf/mcmc-convergence`.
Once the MCMC results are finalised, the analysis will be moved to this appendix and expanded upon.
The following draft text and references may be useful in that expanded write-up.

- Improved $\hat R$ statistic from @vehtari2021rank. Recommended only to use the sample if the value is less than 1.05.
- ESS and autocorrelation related to efficiency
- Traceplots helpful for diagnosis of problems
- Pairs density plots helpful for understanding relationships between parameters, including possible identifiability issues
- Energy plot from @betancourt2017conceptual

@@

1. Calculate the mode, Hessian at the mode, and lower Cholesky
\begin{align*}
\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\texttt{LA}(\btheta, \y)}, \\
\hat{\mathbf{H}} &= - \frac{\partial^2}{\partial \btheta \partial \btheta^\top} \log \tilde p_\texttt{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
\hat{\mathbf{H}}^{-1} &= \hat{\mathbf{L}} \hat{\mathbf{L}}^\top,
\end{align*}
of the Laplace approximation
$$
\tilde p_\texttt{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
$$
where $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
\begin{align*}
\hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \\
\hat{\mathbf{H}}(\btheta) &= - \frac{\partial^2}{\partial \x \partial \x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
\end{align*}
2. Generate a set of nodes $\bu \in \mathcal{Q}(m, k)$ and weights $\omega: \bu \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\bu) = \hat{\btheta} + \mathbf{L} \bu$.
If possible $k \geq 3$ is preferred, though the number of grid points scales exponentially with choice of $k$.
Then, use this quadrature rule to calculate the normalising constant $\tilde p_{\texttt{AQ}}(\y)$ as follows
\begin{equation}
\tilde p_{\texttt{AQ}}(\y) = \sum_{\bu \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\btheta(\bu), \y) \omega(\bu). \label{eq:evidence1}
\end{equation}
4. For $i \in [n]$ generate $l$ nodes $x_i(\bv)$ via a Gauss-Hermite quadrature rule $\bv \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\hat{\mathbf{H}}(\btheta)^{-1}]_i}$ of the Gaussian marginal. A value of $l \geq 4$ is recommended to enable B-spline interpolation.
Then, for $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\bu) \}_{\bu \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
\begin{align*}
\hat \x_{-i}(x_i, \btheta) &= \argmax_{\x_{-i}} \log p(\y, x_i, \x_{-i}, \btheta), \\
\hat{\mathbf{H}}_{-i, -i}(x_i, \btheta) &= - \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
\end{align*}
where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ is initialised at $\hat \x(\btheta)_{-i}$.
6. For $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$ calculate
\begin{equation}
\tilde p_\texttt{AQ}(x_i \, | \, \y) = \frac{\tilde p_\texttt{LA}(x_i, \y)}{\tilde p_{\texttt{AQ}}(\y)}. \label{eq:laplace-marginal}
\end{equation}
where
$$
\tilde p_\texttt{LA}(x_i, \y) = \sum_{\bu \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(x_i, \btheta(\bu), \y) \omega(\bu).
$$
and
$$
\tilde p_\texttt{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\texttt{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
$$
Although Equation \ref{eq:laplace-marginal} can be calculated using the estimate of the evidence in Equation \ref{eq:evidence1} it is more numerically accurate to use the estimate
$$
\tilde p_{\texttt{AQ}}(\y) = \sum_{\bv \in \mathcal{Q}(1, l)} \tilde p_\texttt{LA}(x_i(\bv), \y) \omega(\bv)
$$

7. Given $\{x_i(\bv), \tilde p_\texttt{AQ}(x_i(\bv) \, | \, \y)\}_{\bv \in \mathcal{Q}(1, l)}$ create a spline interpolant to each posterior marginal on the log-scale. Samples, and thereby relevant posterior marginal summaries, may be obtained using inverse transform sampling.

```{r algorithm-flowchart, fig.cap="Flowchart describing the algorithm we propose (work in progress)."}
knitr::include_graphics("depends/algorithm-flowchart.pdf")
```

>>>>

# Simplified Naomi \label{sec:math}

@@

## Implementation

The C++ `TMB` code for the negative log-posterior of the simplified Naomi model is available from the GitHub repository `athowes/elgm-inf`.
For ease of understanding, Table \ref{tab:tmb-lookup} provides correspondence between the mathematical notation used in Section \ref{sec:math} and the variable names used in the `TMB` code, for all hyperparameters and latent field parameters.
For further reference on the `TMB` software see @tmbdocumentation.

@@

<!-- \newpage -->

<!-- ```{cpp, echo=TRUE, eval=FALSE, output.var="ex", code=readLines("naomi_simple.cpp")} -->
<!-- ``` -->

@@

We assessed MCMC convergence and suitability using a range of graphical and numerical tests.
All potential scale reduction factor $\hat R$ statistics [@vehtari2021rank] were below 1.05 and therefore acceptable (Figure \ref{fig:rhat}).
However, even thinning by a factor of 20, samples were not obtained very efficiently, resulting in the majority of effective sample size (ESS) ratios being below 0.5, with some as low as 0.1 (Figure \ref{fig:ratio}).
As a result, the number of obtained ESS varied substantially by parameter (Figure \ref{fig:ess}).
There were no divergent transitions.
Could add energy plot from @betancourt2017conceptual here.

```{r}
np <- readRDS("depends/nuts-params.rds")

n_divergent <- np %>%
  filter(Parameter == "divergent__") %>%
  summarise(n_divergent = sum(Value)) %>%
  pull(n_divergent)

stopifnot(n_divergent == 0)
```

```{r rhat, fig.cap="The potential scale reduction factor compared between- and within- estimates of univariate parameters. It is recommended against to use the results if the value is less than 1.05."}
knitr::include_graphics("depends/rhat.png")
```

```{r ratio, fig.cap="The ESS ratio measures how efficient sampling is."}
knitr::include_graphics("depends/ratio.png")
```

```{r ess, fig.cap="The effective number of samples varied substantially."}
knitr::include_graphics("depends/ess.png")
```

@@

In earlier work, we used a full Laplace approximation to the latent field marginals.
The algorithm for obtaining this approximation is detailed here.

1. Calculate the mode, Hessian at the mode, lower Cholesky, and Laplace approximation
\begin{align*}
\hat{\btheta} &= \argmax_{\btheta} {\tilde p_\texttt{LA}(\btheta, \y)}, \\
\hat{\mathbf{H}} &= - \frac{\partial^2}{\partial \btheta \partial \btheta^\top} \log \tilde p_\texttt{LA}(\btheta, \y) \rvert_{\btheta = \hat \btheta} \\
\hat{\mathbf{H}}^{-1} &= \hat{\mathbf{L}} \hat{\mathbf{L}}^\top, \\
\tilde p_\texttt{LA}(\btheta, \y) &= \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
\end{align*}
where $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \hat{\mathbf{H}}(\btheta)^{-1})$ is a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
\begin{align*}
\hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \\
\hat{\mathbf{H}}(\btheta) &= - \frac{\partial^2}{\partial \x \partial \x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
\end{align*}
2. Generate a set of nodes $\bu \in \mathcal{Q}(m, k)$ and weights $\omega: \bu \to \mathbb{R}$ from a Gauss-Hermite quadrature rule with $k$ nodes per dimension, which are then adapted based on the mode and lower Choleksy via $\btheta(\bu) = \hat{\btheta} + \mathbf{L} \bu$.
Then, use this quadrature rule to calculate the normalising constant $\tilde p_{\texttt{AQ}}(\y)$ as follows
\begin{equation}
\tilde p_{\texttt{AQ}}(\y) = \sum_{\bu \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\btheta(\bu), \y) \omega(\bu). \label{eq:evidence1}
\end{equation}
3. For $i \in [n]$ generate $l$ nodes $x_i(\bv)$ via a Gauss-Hermite quadrature rule $\bv \in \mathcal{Q}(1, l)$ adapted based on the mode $\hat \x(\btheta)_i$ and standard deviation $\sqrt{\text{diag}[\hat{\mathbf{H}}(\btheta)^{-1}]_i}$ of the Gaussian marginal.
$l \geq 4$ is recommended to enable B-spline interpolation.
Then, for $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$  and $\btheta \in \{ \btheta(\bu) \}_{\bu \in \mathcal{Q}(m, k)}$ calculate the modes and Hessians
\begin{align*}
\hat \x_{-i}(x_i, \btheta) &= \argmax_{\x_{-i}} \log p(\y, x_i, \x_{-i}, \btheta), \\
\hat{\mathbf{H}}_{-i, -i}(x_i, \btheta) &= - \frac{\partial^2}{\partial \x_{-i} \partial \x_{-i}^\top} \log p(\y, x_i, \x_{-i}, \btheta) \rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)},
\end{align*}
where optimisation to obtain $\hat \x_{-i}(x_i, \btheta)$ is initialised at $\hat \x(\btheta)_{-i}$.
4. For $x_i \in \{ x_i(\bv) \}_{\bv \in \mathcal{Q}(1, l)}$ calculate
\begin{equation}
\tilde p_\texttt{AQ}(x_i \, | \, \y) = \frac{\tilde p_\texttt{LA}(x_i, \y)}{\tilde p_{\texttt{AQ}}(\y)}. \label{eq:laplace-marginal}
\end{equation}
where
$$
\tilde p_\texttt{LA}(x_i, \y) = \sum_{\bu \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(x_i, \btheta(\bu), \y) \omega(\bu).
$$
and
$$
\tilde p_\texttt{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\texttt{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat \x_{-i}(x_i, \btheta)}.
$$
Although Equation \ref{eq:laplace-marginal} can be calculated using the estimate of the evidence in Equation \ref{eq:evidence1} it is more numerically accurate to use the estimate
$$
\tilde p_{\texttt{AQ}}(\y) = \sum_{\bv \in \mathcal{Q}(1, l)} \tilde p_\texttt{LA}(x_i(\bv), \y) \omega(\bv)
$$

5. Given $\{x_i(\bv), \tilde p_\texttt{AQ}(x_i(\bv) \, | \, \y)\}_{\bv \in \mathcal{Q}(1, l)}$ create a spline interpolant to each posterior marginal on the log-scale. Samples, and thereby relevant posterior marginal summaries, may be obtained using inverse transform sampling.

<!-- ```{r algorithm-flowchart, fig.cap="Flowchart describing the algorithm we propose (work in progress)."} -->
<!-- knitr::include_graphics("depends/algorithm-flowchart.pdf") -->
<!-- ``` -->

### 2023-06-02 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Small edits to MCMC convergence
Add MCMC convergence asssessment to appendix #48

This MCMC took `r out$time` to run

```{r}
bayesplot::color_scheme_set("viridisA")
ggplot2::theme_set(theme_minimal())
```

@@

```{r}
rhats <- bayesplot::rhat(mcmc)
bayesplot::mcmc_rhat(rhats)
(big_rhats <- rhats[rhats > 1.1])
length(big_rhats) / length(rhats)
```

@@

```{r}
ratios <- bayesplot::neff_ratio(mcmc)
bayesplot::mcmc_neff(ratios)
```

@@

data.frame(mcmc_summary) %>%
  tibble::rownames_to_column("param") %>%
  ggplot(aes(x = n_eff)) +
    geom_histogram(alpha = 0.8) +
    labs(x = "ESS", y = "Count")
```

@@

```{r}
np <- bayesplot::nuts_params(mcmc)
```

@@

>>>>

This MCMC took `r round(out$time, 3)` days to run

```{r}
cbpalette <- c("#56B4E9", "#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")

bayesplot::color_scheme_set("viridisA")
ggplot2::theme_set(theme_minimal())
```

@@

```{r}
rhats <- bayesplot::rhat(mcmc)

rhat_plot <- mcmc_rhat_data(rhats) %>%
  ggplot(mapping = aes_(x = ~value, y = ~parameter, col = ~description)) +
  geom_segment(mapping = aes_(yend = ~parameter, xend = ifelse(min(data$value) < 1, 1, -Inf)), na.rm = TRUE, alpha = 0.7) +
  scale_color_manual(values = "#E69F00") +
  geom_vline(xintercept = 1.05, linetype = "dashed", col = "grey40") +
  labs(x = "Potential scale reduction factor", y = "NUTS parameter", col = "") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank()
  )

rhat_plot

pdf("rhat.pdf", h = 3, w = 6.25)

rhat_plot

dev.off()

(big_rhats <- rhats[rhats > 1.1])
length(big_rhats) / length(rhats)

```

@@

```{r}
ratios <- bayesplot::neff_ratio(mcmc)

breaks <- c(0, 0.1, 0.25, 0.5, 0.75, 1)

ratio_plot <- mcmc_neff_data(ratios) %>%
  ggplot(mapping = aes_(x = ~value, y = ~parameter, color = ~description)) +
  geom_segment(aes_(yend = ~parameter, xend = -Inf), na.rm = TRUE, alpha = 0.7) +
  scale_color_manual(values = c("#56B4E9", "#009E73", "#E69F00")) +
  geom_vline(xintercept = 0.1, linetype = "dashed", col = "grey40") +
  geom_vline(xintercept = 0.5, linetype = "dashed", col = "grey40") +
  geom_vline(xintercept = 1, linetype = "dashed", col = "grey40") +
  labs(x = "ESS ratio", y = "NUTS parameter", col = "") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank()
  )

ratio_plot

pdf("ratio.pdf", h = 3, w = 6.25)

ratio_plot

dev.off()
```

@@

ess_plot <- data.frame(mcmc_summary) %>%
  tibble::rownames_to_column("param") %>%
  ggplot(aes(x = n_eff)) +
    geom_histogram(alpha = 0.8) +
    labs(x = "ESS", y = "Count")

ess_plot

pdf("ess.pdf", h = 3, w = 6.25)

ess_plot

dev.off()
```

@@

```{r}
np <- bayesplot::nuts_params(mcmc)

saveRDS(np, "nuts-params.rds")
```

@@

# Original computing environment {-}

```{r}
sessionInfo()
```

# Bibliography {-}
